Open source landscape ecology tools \thanks{Grants or other notes about the article that should go on the front page should be placed here. General acknowledgments should be placed at the end of the article.}

### 2020-12-15 paper/paper.tex
:pencil: Adding  packages addressing some comments section of survey

Since most \emph{R} packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
There are two major online platforms to host \emph{R} packages and make them accessible to potential users.
Firstly, \emph{GitHub} allows to host source code under version control and users can install packages with one line using the \emph{remotes} \cite{Hester2020} package.
Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
\cite{Wickham2015}.
For developers, the guaranteed technical quality standard on \emph{CRAN} requires more initial work for compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e.~package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
For more information about \emph{R} package development, see \cite{Wickham2015}.

@@

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The probably most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
Alternatively, the \emph{stars} package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.

@@

Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a \emph{R} session.
Since public available data is becoming more and more prominent, so are \emph{R} packages to accesses them.
Packages include the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA, the \emph{elevatr} package to access raster elevation data based on the Amazon Web Services Terrain Tiles \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, \emph{rnaturalearth} to access the Natural Earth database \cite{South2017}, or \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}.
Furthermore, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4, while the new in-development package \emph{geodata} provides an access to data from WorldClim v2.
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Creating maps is essential when working with spatial data.
It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\emph{R} allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.

@@

\hypertarget{sec:landscape_metrics}{%
\paragraph{Landscape metrics}\label{sec:landscape_metrics}}

@@

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020} and \emph{PresenceAbsence} \cite{Freeman2008} packages.

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.~amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011} landscapes are described by nodes (i.e.~habitat patches) connected by and functional connections called links \cite{Laita2011}.
The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More general, the \emph{igraph} packages \cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are another tool to model functional connectivity.
The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, or \emph{HierDpart} packages \cite{Qin2019}.

@@

@@

To better understand how \emph{R} is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
Other, less frequent answers, were ``Data scientists'', ``None of the above'', ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity'', followed by ``land use management'', ``landscape connectivity'' and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data'', ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate'', ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig. \ref{fig:fig-survey} D)).

@@

\FloatBarrier

\hypertarget{missing-packages}{%
\section{Missing packages}\label{missing-packages}}

@@

Since its first introduction in 1995, \emph{R} has come a long way from a statistical programming language to a powerful tool for landscape ecology.
Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
However, since the scientific field of landscape ecology constantly changes and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.

>>>>

Since most \emph{R} packages are developed and maintained by the community, this shows how open-source software development can facilitate innovation, reproducibility and reuse of code.
There are two major online platforms to host \emph{R} packages and make them accessible to potential users.
Firstly, \emph{GitHub} allows to host source code under version control and users can install packages with one line using the \emph{remotes} \cite{Hester2020} package.
Additionally, hosting the package on \emph{GitHub} provides many useful features to collaborate and communicate between developers as well as users \cite{Wickham2015}, or integrated unit testing.
Secondly, the \emph{Comprehensive R Archive Network} (\emph{CRAN}) provides a large visibility to the community, ease of installation and a technical quality standard, including checks for common problems on all major operating systems.
\cite{Wickham2015}.
For developers, the guaranteed technical quality standard on \emph{CRAN} requires more initial work for compared to \emph{GitHub}, while it ensures for users that the package can be installed on their machine.
Additionally, the technical quality standard on \emph{CRAN} also facilitates reproducibility and reuse of code, as shown by many reverse dependencies of \emph{R} packages, i.e.~package \emph{x} requires and uses code from package \emph{y} \cite{vonKrogh2006,St.Laurent2008,Steiniger2009,Prlic2012,Powers2019}.
Thus, most packages can be found on both platforms and many developers use \emph{GitHub} for daily developing and \emph{CRAN} to publish stable releases of the packages.
For more information about \emph{R} package development, see \cite{Wickham2015}.

@@

In the raster data model, surfaces are divided into cells, where each cell stores a numeric value.
The values could represent discrete phenomena, such as a class number of a land cover category, or continuous phenomena, such as elevation values.
The probably most prominent package allowing for raster data representation is \emph{raster} \cite{Hijmans2019}.
A \emph{raster} successor aimed at the simpler interface and improved performance, \emph{terra}, is now being developed \cite{R-terra}, however, it could take several years for this package to be adopted by users.
Alternatively, the \emph{stars} package can be used to read and process raster data with a focus on spatial-temporal data cubes \cite{Pebesma2019}.
Additionally, there are packages that improve some basics raster operations in terms of computational performance or compatibility between raster and vector operations, such as \emph{fasterize} \cite{Ross2020}, \emph{rasterDT} \cite{OBrien2020}, or \emph{exactextractr} \cite{Baston2020}.

@@

Nowadays, spatial data for many spatial scales is available from an abundance of online accessible sources.
A lot of this data is publicly available, either as a direct download or through an API connection, and this can be used by several packages to download the spatial data directly into a \emph{R} session.
Since public available data is becoming more and more prominent, so are \emph{R} packages to accesses them. Packages include \emph{rnaturalearth} \cite{South2017} to access the Natural Earth database, the \emph{elevatr} package to access raster elevation data \cite{R-elevatr}, the \emph{rgbif} to access the Global Biodiversity Information Facility (GBIF) portal \cite{R-rgbif}, the \emph{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \emph{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \emph{FedData} package \cite{R-FedData} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \emph{getlandsat} package \cite{Chamberlain2018} allows to download data from the Landsat 8, while the \emph{MODIS} package \cite{Mattiuzzi2020} allows to download MODIS products.
Also, the \texttt{get\_data()} function from the \emph{raster} package allows to download climatic and bioclimatic data from WorldClim v1.4 (for WorldClim v2, see the in-development package \emph{geodata} ).
Additionally, the \emph{rgee} package \cite{R-rgee} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

Creating maps is essential when working with spatial data.
It has a role in checking the spatial and value-related quality of data, data exploration, and finally communicating results.
\emph{R} allows to create two major types of maps.
Firstly, static maps in which the developer has full control over the presentation of the map and secondly, interactive maps in which the user has the possibility to modify the map by e.g.~changing the displayed values.
All packages listed in Section \ref{sec:spatrep} have build-in methods for plotting spatial data using the generic \texttt{plot()} function.
However, the generic functions are focused on quick visual inspection of the data, rather than the creating complete maps.
Advanced plotting methods for raster objects can be found in the \emph{rasterVis} package \cite{Lamigueiro2020}.
Alternatively, the \emph{mapview} package can be used for quick interactive visualization of spatial data \cite{R-mapview}.
The \emph{tmap} package provides a coherent plotting system for static and interactive maps that is based on the layered grammar of graphics \cite{Tennekes2018} and aims for creating publication-ready maps.
Both, \emph{tmap} and \emph{mapview} build upon the \emph{leaflet} package and leaflet javascript library \cite{R-leaflet}.
Thematic maps, including besides others proportional symbols, chroropleth, or typology, can be created with the \emph{cartography} package \cite{Giraud2016}.
Also the popular plotting package \emph{ggplot2} \cite{Wickham2016a} has an extension especially designed for plotting spatial data named \emph{ggspatial} \cite{R-ggspatial}.

A slightly different approach of visualizing spatial data is the \emph{rayshader} package \cite{Morgen-Wall2020} which allows to create topographical 2D and 3D maps.

@@

\hypertarget{sec:landscape_metrics}{%
\paragraph{Quantify landscape characteristics}\label{sec:landscape_metrics}}

@@

More recently, surface metrics were suggested as an alternative to landscape metrics for continuous raster data \cite{McGarigal2009}. The \emph{geodiv} package \cite{Smith2020} allows to gradient surface metrics to characterize continuous analysis of landscape features.

@@

Because the used modeling approaches are diverse \cite{Hooten2011,Kerr2011,Fletcher2019}, there is also a large number of \emph{R} packages used for SDMs.
Popular approaches and packages include generalized linear models using, e.g., the \emph{stats} \cite{RCoreTeam2019} package; generalized additive models using, e.g., the \emph{mgcv} \cite{Wood2017} or \emph{lme4} \cite{Bates2015} package; classification and regression trees (CART) using, e.g., the \emph{rpart} \cite{Therneau2019}, \emph{randomForest} \cite{Liaw2002} or \emph{ranger} \cite{Wright2017} package or multivariate data analysis using , e.g., the \emph{ade4} \cite{Dray2007} or \emph{vegan} \cite{Oksanen2019} package.
Of course, also packages specifically designed for SDM exists.
This includes the \emph{dismo} \cite{Hijmans2017}, \emph{sdm} \cite{Naimi2016}, \emph{ecospat} \cite{Broennimann2020}, \emph{biomod2} \cite{Thuiller2020}, \emph{PresenceAbsence} \cite{Freeman2008} or \emph{zoon} \cite{Golding2018} packages.
Related to SDM, there are several packages to estimate home ranges and habitat selection, including \emph{adehabitatHR} and \emph{adehabitatHS} \cite{Calenge2006} or \emph{amt \cite{Signer2019}}

@@

Connectivity is one of the core elements of landscape structure \cite{Taylor1993} and thus one of the core concepts of landscape ecology \cite{With2019}.
Landscape connectivity describes how landscape characteristics facilitate or hinder the movement of species \cite{Tischendorf2000} or other aspects of mobility, such as dispersal, gene or nutrient flow \cite{With2019}.
While structural connectivity focuses only on landscape characteristics (e.g., movement corridors, barriers), functional connectivity also includes behavior characteristics of the species \cite{Tischendorf2000,With2019}.
Given its wide concept, many different measures of connectivity exists \cite{Kindlmann2008}.
On patch level structural connectivity can be measured using nearest-neighbor distances or characterizations of the patch neighborhood (e.g.~amount of suitable habitat) \cite{Kindlmann2008,With2019}.
Such measures are provided within the \emph{landscapemetrics} package (see \ref{sec:landscape_metrics}).
Furthermore, the \emph{lconnect} package \cite{Mestre2019} and \emph{Makurhini} package \cite{Godinez-Gomez2020} provide several landscape connectivity metrics.
Another way to describe connectivity is based on graph theory with the advantages that also functional connectivity can be included \cite{Kindlmann2008}.
In graph theory \cite{Laita2011} landscapes are described by nodes (i.e.~habitat patches) connected by and functional connections called links \cite{Laita2011}.
The \emph{grainscape} package \cite{Chubaty2020} provides a tool to model connectivity based on spatially explicit networks.
More general, the \emph{igraph} packages \cite{Csardi2006} provides functionality related to graph theory.
Resistance surfaces and least-cost paths are another tool to model functional connectivity.
The resistance surface describes effects of facilitating or hindering characteristics of the landscape for an organism moving within it \cite{Adriaensen2003}.
Least-cost paths can be calculated using the \emph{gdistance} package \cite{vanEtten2017}.
Absorbing Markov chains that combines random walk theory and mortality based on the landscape characteristics \cite{Fletcher2019a} and is provided by the recently published \emph{samc} package \cite{Marx2020}.

@@

Since describing connectivity between two locations is one of the fundamental steps of landscape genetics, all packages useful for connectivity (see \ref{sec:landscape_genetics}) are also of importance for landscape genetics.
Further functionality for landscape genetics can be found in the \emph{graphs4lg} \cite{Savary2020}, \emph{PopGenReport} \cite{Adamack2014,Gruber2015}, \emph{HierDpart} packages \cite{Qin2019}, or \emph{GeNetIt} \cite{Murphy2010}.

@@

\hypertarget{various}{%
\paragraph{Various}\label{various}}

@@

To better understand how \emph{R} is used by the landscape ecology community, we conducted a short survey and used mailing lists and social media to reach the community.
In total, the survey was answered by 103 participants, of which the the majority were either ``PhD students'' (34\%), followed by ``Post-Docs'' (28.2\%) and ``Professors'' (12.6\%).
Other, less frequent answers, were ``Data scientists,'' ``None of the above,'' ``Government employees'' ``Master's degree student'' and ``Bachelor's degree student'' (in decreasing order).

@@

We asked the participants to select which terms describe their research topics the best and options that were selected by more than 10\% of participants included ``biodiversity,'' followed by ``land use management,'' ``landscape connectivity'' and ``nature conservation'' (Fig. \ref{fig:fig-survey} A)).

Next, we were interested in the most important tasks to the workflow of the participants.
Not surprisingly, ``(pre-)processing of data,'' ``spatial statistics'' and ``creating maps'' were the most selected options (Fig. \ref{fig:fig-survey} B)).
Interestingly, the available options seemed to describe the most important task to the workflow quite well since only very few participants selected the ``others'' option (all options with less than five total answers were classified as ``others'').

@@

Lastly, when asked how use \emph{R} us currently for landscape ecology, the vast majority of participants answered with either ``very useful'' or ``useful'' (summarized 91.26\%) and only only very few participants evaluated \emph{R} as ``intermediate,'' ``not useful'' or ``not useful at all'' (summarized 8.74\%; Fig. \ref{fig:fig-survey} D)).

@@

@@

Since its first introduction in 1995, \emph{R} has come a long way from an exclusively statistical programming language to a powerful tool for landscape ecology.
Today, many \emph{R} packages, mainly developed by the community itself, provide a vast collection of functions and algorithms aimed for spatial data handling and analysis.
The highly dynamic development of \emph{R} packages for landscape ecology also shows the strength of open-source software with its high innovation, transparency, reliability, and longevity.
However, since the scientific field of landscape ecology constantly develops and improves, also the \emph{R} programming language and its packages need to change and adapt these changes.

\hypertarget{acknowledgments}{%
\subsection{Acknowledgments}\label{acknowledgments}}

\hypertarget{conflict-of-interest}{%
\paragraph{Conflict of interest}\label{conflict-of-interest}}

MHKH and JN are authors of the \emph{landscapemetrics} and \emph{landscapetools} package.
JN is author of \emph{belg} and \emph{motif} package.

\hypertarget{funding}{%
\paragraph{Funding}\label{funding}}

MHKH was supported by the German Research Association (DFG) Research Training Group 1644 `Scaling Problems in Statistics,' grant number 152112243.

\hypertarget{author-contributions}{%
\paragraph{Author contributions}\label{author-contributions}}

MHKH and JN designed the survey form and analyzed the responses of the participants.
MHKH and JN drafted the manuscript and all authors contributed critically to the manuscript and gave final approval for publication.
We used the `sequence--determines--credit' approach (SDC) for the sequence of authors.

\hypertarget{human-and-animal-rights-and-informed-consent}{%
\paragraph{Human and Animal Rights and Informed Consent}\label{human-and-animal-rights-and-informed-consent}}

This article does not contain any studies with human or animal subjects performed by any of the authors.
